The Supervisionary proof-checking kernel

### 2021-09-30 paper/prisc-2022.tex
prisc-2022: some more minor rewriting, tightening, etc.

Though ostensibly wildly different, in \emph{some} respects the goals of operating system and proof-checking kernels are the same: the maintenance of important system invariants in the face of unbounded interaction with untrusted code.
The mechanisms employed to protect these invariants, by each system, are different however.
Operating systems kernels are cosseted by paged memory protection and execute at a relative level of privilege compared to untrusted code.
On the other hand, LCF-style proof-checking kernels are typically implemented as libraries within a \emph{meta-language}---usually an ML dialect---with invariants maintained by type abstraction and module encapsulation.
Logical soundness is ultimately guaranteed by type-soundness of the meta-language.

What happens if we structure proof-checking software like \emph{supervisory} system software---such as operating systems, hypervisors, and programming language runtimes?
We are currently developing a prototype system, \emph{Supervisionary}, implemented as a WebAssembly host, to explore the consequences of this design.
In one sense, Supervisionary can be used as a standard proof-checker for Gordon's HOL, driven by an untrusted WebAssembly program.
However, in another sense, Supervisionary exposes a novel abstract machine to untrusted code, in much the same way that the Linux kernel exposes an abstract machine to user-space code through its system call interface, for example.
We pr\`{e}cis some of the consequences of Supervisionary's design below, and argue that this novel abstract machine is of interest.

\paragraph{No meta-language} One immediate consequence of Supervisionary's design is the abolition of the LCF meta-language.
Rather, Supervisionary's kernel is implemented in Rust.
On the other hand, untrusted code which ``drives'' the Supervisionary kernel to produce an authenticated theorem may be written in any language that can be compiled to code that is binary-compatible with the Supervisionary abstract machine, or even a mixture of them, in much the same way that Linux user-space software can be written in Haskell, Ada, or any other language that can be linked together, despite the kernel being written in C.

\paragraph{ABI design} Rust is a notably \emph{unsafe} systems programming language, yet its use in the Supervisionary kernel does not pose any risk to the logical soundness of the system, providing we expose a carefully designed ABI.
The Supervisionary kernel manages various private memory spaces, or heaps, within which \emph{kernel objects} are managed.
These kernel objects correspond to the parephenalia of a typical HOL implementation---type-formers, types, term constants, terms, and theorems---but are never directly exposed to untrusted code.
Rather, kernel objects are allocated by the kernel in one of its private heaps, through ABI functions like the following:
\begin{displaymath}
\mathtt{Term.Handle.AllocateApplication(left, right, out)}
\end{displaymath}
Above, both $\mathtt{left}$ and $\mathtt{right}$ passed to the function are kernel handles, assumed to point-to allocated terms.
On the other hand, the $\mathtt{out}$ pointer points-into the heap of the code calling the function (note that the kernel can freely read from, and write to, the memory of untrusted code, given it is supervisory software).
Providing both $\mathtt{left}$ and $\mathtt{right}$ indeed point-to allocated terms, and their types match, a new kernel object corresponding to a term application is allocated, with internal pointers to the functional- and argument-terms, in the kernel's heap.
This new object is pointed-to by a freshly-generated handle, returned back to the caller via the $\mathtt{out}$ pointer.

>>>>

Though ostensibly wildly different, in \emph{some} respects the goals of operating system and proof-checking kernels are the same: the maintenance of important system invariants in the face of unbounded interaction with untrusted code.
The mechanisms employed to protect these invariants, by each system, are different however.
Operating systems kernels are cosseted by paged memory protection and execute at a relative level of privilege compared to untrusted code, which interacts with the kernel through a binary interface.
On the other hand, LCF-style proof-checking kernels are typically implemented as libraries within a \emph{meta-language}---usually an ML dialect---with invariants maintained by type abstraction and module encapsulation, and linked against untrusted code typically written in the meta-language.

What happens if we structure proof-checking software like \emph{supervisory} system software---such as operating systems, hypervisors, and programming language runtimes?
We are currently developing a prototype system, \emph{Supervisionary}, implemented as a WebAssembly host, to explore the consequences of this design.
In one sense, Supervisionary can be used as a standard proof-checker for Gordon's HOL, driven by an untrusted WebAssembly program that interacts with the kernel via a binary interface which defines an abstract machine, in much the same way that the Linux kernel exposes an abstract machine to user-space code through its system call interface, for example.
However, in another sense, Supervisionary can be generalised to a programming environment that can host arbitrary programs.
We briefly describe some of the consequences of Supervisionary's design below, and argue that this novel abstract machine is of interest.

\paragraph{No meta-language} One immediate consequence of Supervisionary's design is the lack of any meta-language.
Rather, Supervisionary's kernel is implemented in Rust.
On the other hand, untrusted code which ``drives'' the Supervisionary kernel to produce an authenticated theorem may be written in any language that can be compiled to code that is binary-compatible with the Supervisionary abstract machine, or even a mixture of them, in much the same way that Linux user-space software can be written in Haskell, Ada, or any other language that can be linked together, despite the kernel being written in C.

\paragraph{ABI design} Rust is a notably \emph{unsafe} systems programming language, yet its use in the Supervisionary kernel does not pose any risk to the logical soundness of the system, providing we expose a carefully designed ABI.
The Supervisionary kernel manages various private memory spaces, or heaps, within which \emph{kernel objects} are managed.
These kernel objects correspond to the parephenalia of a typical HOL implementation---type-formers, types, term constants, terms, and theorems---but are never directly exposed to untrusted code.
Rather, kernel objects are allocated by the kernel in one of its private heaps, through ABI functions like the following:
\begin{displaymath}
\mathtt{Term.Handle.AllocateApplication(left, right, out)}
\end{displaymath}
Here, $\mathtt{left}$ and $\mathtt{right}$ are kernel handles, assumed to point-to allocated terms in the kernel's heaps.
On the other hand, the $\mathtt{out}$ pointer points-into the heap of the code calling the function---the kernel can freely read from, and write to, the memory of untrusted code.
Assuming $\mathtt{left}$ and $\mathtt{right}$ indeed point-to allocated terms, and other checks pass, a new kernel object corresponding to a term application is allocated.
This object contains internal pointers to the functional- and argument-terms, in the kernel's heap.
The kernel generates a fresh handle that points-to this new object, which is returned back to the caller via the $\mathtt{out}$ pointer.
